---
title: "R for Spatial Analysis"
author: "Jackson Voelkel"
date: "November 18th, 2016"
output:
  revealjs::revealjs_presentation:
    transition: slide
    self_contained: false
---



# Introduction
```{r echo=F,warning=F,message=F}
library(knitr)
```

## Me

- Background
- Computer Experience

## What is R?

- Language
- Application
- "Do it all" Workbench

## What can I do in R?

- Advanced (or not so advanced) statistics
- Powerful plotting
- Read almost any file type
- Write almost any file type
- Almost anything you'd ever want to!

## Reasons to use R:

- Programatic
- Reproducible
- Cross Platform: PC, Mac, and Linux!
- Low(ish) level
- Open Source and community driven
- Free
- Free

## Drawbacks to using R:

- Steep learning curve
- Can be incredibly complex for simple tasks
- Not a GUI, and not overtly intuitive

R won't replace spatial analysis platforms like ArcGIS, however it's "language" format will allow for incredibly powerful, complex, and custom geoprocessing tasks to be performed.

## What does R look like?

```{r}
head(mtcars)
```

## Reading in Geospatial Data

Adaptation of the Geospatial Data Abstraction Library (GDAL) is a perfect example of R's incorporation of a wide swath of existing open source libraries. 

## 


##

Just like with a table, this new "SpatialPolygonsDataFrame" is subsetable:

