---
title: 'A Template for HW solutions; M362M, Fall 2020'
author: 'Jane B. Student, eid:jbs123'
date: '2020-09-01'
output: pdf_document
---

<!-- The code chunk below contains some settings that will  -->
<!-- make your R code look better in the output pdf file.  -->
<!-- If you are curious about what each option below does, -->
<!-- go to https://yihui.org/knitr/options/ -->
<!-- If not, feel free to disregard everything ...  -->
```{r echo=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align="center",
  fig.pos="t",
  strip.white = TRUE
)
```
<!-- ... up to here. -->

## Problem 1.1

This is where I would explain what I am doing. 

### Part a.

```{r}
1/238746238746
```
You can write an explanation here, too. 

### Part b.
```{r}
2^45
```

### Part e.
```{r}
cos(pi/8)
```
  
### Part f.
```{r}
exp(2)
```
  
  
### Part h.
```{r}
log10(2)
```

## Problem 1.2

### Part 3.
```{r}
A = matrix(c(1, 2, -1, 3), nrow = 2, byrow = TRUE)
A
```
What would happen if we left out `nrow = 2` or `byrow = TRUE`?

## Problem 1.3.

### Part 2. 
```{r}
my_function <- function(x) { 
  if (2 < x & x < 3) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}
```
Let's see if it works (make sure to evaluate the chunk above first; otherwise, R will not know what `my_function` is):

```{r}
my_function(5)
my_function(2.3)
```

A much more concise function can be written:

```{r}
my_shorter_function <- function(x) {
  return(x > 2 & x < 3)
}
```

```{r}
my_function(5)
my_function(2.3)
```
This works because the expression such as `x>2` in R has a logical (i.e., boolean, i.e. `TRUE` or `FALSE`) value. 
