\newcommand{\PP}{{\mathbb{P}}}
\newcommand{\N}{{\mathbb{N}}}
\newcommand{\sS}{{\mathcal{S}}}
\newcommand{\EE}{{\mathbb{E}}}
\newcommand{\ld}{\lambda}
\newcommand{\eand}{\text{ and }}

# (PART) Probability and Simulation {-}

# Simulation of Random Variables and Monte Carlo

In the spirit of "learn by doing", these lecture notes contain many "Problems".
Those with solutions usually introduce new concepts and feature a *Comments*
section right after the solution. These comments are subdivided into *R* and
*Math* comments focusing on the computational or conceptual features,
respectively. Note that you are not expected to be able to do the solved
problems before reading their solutions and comments, so don't worry if you
cannot. It is a good practice to try, though. Problems which are left unsolved,
however, do not feature any new ideas and are there to help you practice the
skills presented before. 

## Simulation of some common probability distributions

... where we also review some probability along the way.

```{exercise}
''Draw'' 50 simulations from the geometric distribution with parameter $p=0.4$.
```
**Solution:** 
```{r}
rgeom(50,prob=0.4)
```
**Comments**

*R:* R makes it very easy to simulate draws from a large class of *named
distributions*^[There are infinitely many ways random variables can be
distributed. Indeed, in the discrete ${\mathbb N}$-valued case only, any
sequence of nonnegative numbers $(p_n)_n$ such that $\sum_n p_n=1$ defines
*a* probability distribution. It turns out, however, that a small-ish number of
distributions appear in nature much more often then the rest. These
distributions, like the normal, uniform, exponential, binomial, etc. turn out to
be so important that they each get a name (hence *named distributions*). ],
such as geometric, binomial, uniform, normal, etc. For a list of all available
distributions, run `help("distributions")` Each available distribution has an *R
name*; the uniform is `unif` the normal is `norm` and the binomial is `binom`,
etc. If you want to simulate $n$ draws (aka a *sample* of size $n$) from a
distribution, you form a full command by appending the letter `r` to its R name
and use $n$ as an argument. That is how we arrived to `rgeom(50)` in the
solution above. The additional arguments of the function `rgeom` have to do with
the parameters of that distribution. Which parameters go with which
distributions, and how to input them as arguments to `rgeom` or `rnorm` is best
looked up in R's extensive documentation. Try `help("rnorm")`, for example. 

*Math:* 
You could spend your whole life trying to understand what it really means to
"simulate" or "generate" a random number. The numbers you obtain from so-called
*random number generators* (RNG) are never random. In fact, they are completely
deterministically generated. Still, sequences of numbers obtained from (good)
random number generators share so many properties with sequences of mythical
*truly* random numbers, that we can use them as if they were truly random. For
the purposes of this class, you can assume that the numbers R gives you as
*random* are random enough. Random number generation is a fascinating topic at
the intersection of number theory, probability, statistics, computer science and
even philosophy, but we do not have the time to cover any of it in this class.
If you want to read a story about a particularly bad random number generator, go
[here](https://en.wikipedia.org/wiki/RANDU).



You might have encountered a geometric distribution before. A random variable with that
distribution can take any positive integer value or $0$, i.e., its support is 
${\mathbb N}_0=\{0,1,2,3,\dots\}$. 
As you can see from the output above, the value $0$ appears more often than the value $3$,
and the value $23$ does not appear at all in this particular simulation run. The 
probability of seeing the value $k\in \{0,1,2,3,\dots\}$ as a result of a single 
draw is given by $(1-p)^k p$, where $p$ is called the *parameter* of the distribution. 

That corresponds to the following interpretation of the geometric distribution:
keep tossing a biased coin (with probability p of obtaining H) until you see the first H; the  number Ts before that is that value your geometric random variable^[Some books will define the geometric random variables as the number of *tosses* (and not Ts) before the first H is obtained. In that case, the final H is included into the count. Clearly, this definition and the one we have given differ by $1$, and this is really not a big deal, but you have to be careful about what is meant when a geometric random variable is mentioned.]
If we put these probabilities in a single table (and choose $p=0.4$, for example) it is 
going to look like this:
```{r echo=FALSE}
k=0:7
df = data.frame(dgeom(k, prob=0.4))
names(df)=c("Prob.")
tf=data.frame(t(df))
tf$elip = c("...")
names(tf) = c(0:7,'...')
knitr::kable(tf, digits=3)
```

Of course, the possible values our random variable can take do not stop at $7$.
In fact, there are infinitely many possible values, but we do not have infinite
space. Note that even though the value $23$ does not appear in the output of the
command `rgeom` above, it probably would if we simulated many more than $50$
values. Let's try it with $500$ draws - the table below counts how many $0s$,
$1s$, $2s$, etc. we got:

```{r echo=FALSE}
X = rgeom(500, prob = 0.4)
knitr::kable(t(table(X)))
```

Still no luck, but we do observe values above 5 more often. By trial and error,
we arrive at about $1,000,000$ as the required number of simulations:

```{r echo=FALSE, message=FALSE}
library(tidyverse)
cts = as.vector(table(rgeom(1000000, prob = 0.4)))
out  = as_tibble(t(cts), .name_repair = "unique")
names(out) = as.character(c(-1+1:length(out)))
out = select(out, head(names(out),4), tail(names(out),5) )
out[,5]=c("...")
names(out)[5]="..." 
kable(out)
```


```{exercise}
Compute the probability that among $1,000,000$ draws of a geometric random
variable with parameter $p=0.4$, we never see a number greater than $22$.
```

**Solution:** First, we compute the probability that the value seen in a *single* draw does not exceed $22$:
```{r echo=2}
options(digits=7)
pgeom(22,prob=0.4)
```

Different draws are *independent* of each other, so we need to raise this to the power $1,000,000$. 
```{r}
(pgeom(22,prob=0.4))^(1000000)
```
**Comments:**

*R.* The command we used here is `pgeom` which is a cousin of `rgeom`. In
general, R commands that involve named probability distributions consist of two
parts. The prefix, i.e., the initial letter (`p` in this case) stands for the
operation you want to perform, and the rest is the R name of the distribution.
There are 4 prefixes, and the commands they produce are

| Prefix   | Description                                    |
|:----------|:------------------------------------------------|
| `r` | Simulate **r**andom draws from the distribution.  |
| `p` | Compute the cumulative **p**robability distribution function (cdf) (**NOT pdf**)|
| `d` | Compute the probability **d**ensity (pdf) or the probability mass function (pmf) 
| `q` | Compute the **q**uantile function |

(see the Math section below for the reminder of what these things are). In this
problem, we are dealing with a geometric random variable $X$, which has a
discrete distribution with support $0,1,2,3,\dots$. Therefore, the R name is
`geom`. We are interested in the probability $\PP[ X\leq 22]$, which
corresponds to the cdf of $X$ at $x=22$, so we use the 
the prefix `p`. Finally, we used the named parameter `p` and gave it the value `p = 0.4`, because the geometric distribution has a single parameter $p$.

This problem also gives us a chance to discuss precision. As you can see, the
probability of a single draw not exceeding $22$ is very close to $1$. In fact,
it is equal to it to 5 decimal places. By default, R displays 7 significant
digits of a number. That is enough for most applications (and barely enough for
this one), but sometimes we need more. For example, let's try to compute the
probability of seeing no T (tails) in 10 tosses of a biased coin, where the
probability of H (heads) is 0.9.
```{r almost-one}
1-0.1^10
```
While very close to it, this probability is clearly not equal to $1$, as suggested by the output above. 
The culprit is the default precision. We can increase the precision (up to $22$ digits) using the  `options` command
```{r echo = 1:2}
options(digits=17)
1-0.1^10
options(digits=7)
```
Precision issues like this one should not appear in this course, but they will
out there "in the wild", so it might be a good idea to be aware of them.

*Math.* If you forgot all about pdfs, cdfs and such things here is a little reminder:

|   |   |
|---|--------------------|
| cdf | $F(x) = \PP[X\leq x]$|
| pdf | $f(x)$ such that $\PP[X \in [a,b]] = \int_a^b f(x) \, dx$ for all $a<b$ |
| pmf | $p(x)$ such that $\PP[X=a_n] = p(a_n)$ for some sequence $a_n$ |
| qf | $q(p)$ is a number such that $\PP[ X \leq q(p)] = p$


Those random variables that admit a pdf are called **continuous**. The prime
examples are the normal, or the exponential distribution. The ones where a pmf
exists are called **discrete**. The sequence $a_n$ covers all values that such
a, discrete, random variable can take. Most often, $a_n$ either covers the set
of all natural numbers $0,1,2,\dots$ or a finite subset such as $1,2,3,4,5,6$. 

Coming back to our original problem, we note that the probability we obtained is
quite small. Since $1/0.000372$ is about $2690$, we would have to run about
$2690$ rounds of $1,000,000$ simulations before the largest number falls below
$23$.


```{exercise} 
Compute the $0.05$, $0.1$, $0.4$, $0.6$ and $0.95$ quantiles of the normal 
distribution with mean $1$ and standard deviation $2$.
```
**Solution:**
```{r}
qnorm( c(0.05, 0.1, 0.4, 0.6, 0.95), mean = 1, sd = 2)
```

*R.* The function we used is `qnorm`, with the prefix `q` which computes the
quantile function and the R name `norm` because we are looking for the quantiles
of the normal distribution. The additional (named) parameters are where the
parameters of the distribution come in (the mean and the standard variation) in
this case. Note how we plugged in the entire vector
`c(0.05, 0.1, 0.4, 0.6, 0.98)` instead of a single value into `qnorm`. You can
do that because this function is **vectorized**. That means that if you give it
a vector as an argument, it will "apply itself" to each component of the vector
separately, and return the vector of results. Many (but not all) functions in R
are vectorized^[The function `sum` adds up all the components of the vector.
You would not want such a function to be vectorized. If it were, it would return
exactly the same vector it got as input.].

As a sanity check, let's apply `pnrom` (which computes the cdf of the normal) to these quantile values:
```{r}
p =  qnorm( c(0.05, 0.1, 0.4, 0.6, 0.95), mean = 1, sd = 2)
pnorm( p , mean = 1, sd = 2)
```
As expected, we got the original values back - the normal quantile function and its cdf are inverses of each other. 

*Math.* Computing the cdf of a standard normal is the same thing reading a
*normal table*. Computing a quantile is the opposite; you go into the middle of
the table and find your value, and then figure out which "Z" would give you that
value. 


```{exercise}
Simulate $60$ throws of a fair $10$-sided die. 
```
**Solution:**
```{r}
sample( 1:10 , 60, replace = TRUE)
```

**Comments**:

*Math.* Let $X$ denote the outcome of a single throw of a fair $10$-sided die.
The distribution of $X$ is discrete (it can only take the values
$1,2,\dots, 10$) but it is not one of the more famous named distributions. I
guess you could call it a *discrete uniform on ${1,2,\dots, 10}$*, but a better
way to describe such distribution is by a **distribution table**, which is
really just a list of possible values a random variable can take, together with
their, respective, probabilities. In this case, 

```{r echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
options(knitr.table.format = "html") 
library(tidyverse)
x = 1:10
y = rep("0.1", 10)
knitr::kable(t(y), col.names = 1:10, escape = FALSE)

```
*R.* The command used to draw a sample from a (finite) collection is, of, course
`sample`. The first argument is a vector, and it plays the role of the "bag"
from which you are drawing. If we are interested in repeated, random samples, we
also need to specify `replace = FALSE` otherwise, you could draw any single
number at most once:
```{r}
sample(1:10, 8, replace = FALSE)
```
 With more than 10 draws, we would run out of numbers to draw:
```{r error=TRUE}
sample(1:10, 12, replace = FALSE)
```
The bag you draw from can contain objects other than numbers:
```{r echo=2}
set.seed(400)
sample( c("Picard", "Data", "Geordi"), 9, replace = TRUE)
```
So far, each object in the bag had the same probability of being drawn. You can
use the `sample` command to produce a *weighted* sample, too. For example, if we
wanted to simulate $10$ draws from the following distribution
```{r echo=FALSE}
knitr::kable(t(c(0.2,0.7, 0.1)), col.names = c(1,2,3)) %>% 
   kable_styling(full_width = F)
```
we would use the additional argument `prob`:
```{r echo=2}
set.seed(513)
sample( c(1,2,3), 10, replace = TRUE, prob = c(0.2,0.7, 0.1))
```
Note how it is mostly $2$s, as expected.


```{exercise}
Draw a sample of size $n=10$ from $N(1,2)$, i.e., from the normal distribution
with parameters $\mu=1$, $\sigma = 2$. Plot a histogram of the obtained
values. Repeat for $n=100$ and $n=100000$. 
```
**Solution:**
```{r}
x = rnorm(10, mean = 1, sd = 2)
hist(x)
```
```{r}
x = rnorm(100, mean = 1, sd = 2)
hist(x)
```
```{r}
x = rnorm(100000, mean = 1, sd = 2)
hist(x)
```
**Comments:**

*R.* It cannot be simpler! You use the command `hist`, feed it a vector of
values, and it produces a histogram. It will even label the axes for you. If you
want to learn how to tweak various features of your histogram, type `?hist`.

<!-- Esthetically, the built-in histograms leave something to be desired. We can do better, using the package `ggplot2`. You don't have to use it in this class, but if you want to, you install it first by running `install.packages("ggplot2")` (you have to do this only once). Then, every time you want to use it, you run `library(ggplot2)` to notify R that you are about to use a function from that package. It would take a whole semester to learn everything there is to know about `ggplot2`; I will only show what a histogram looks like in it: -->
<!-- ```{r} -->
<!-- library(ggplot2) -->
<!-- x = rnorm(100000, mean = 1, sd = 2) -->
<!-- qplot(x, bins=40) -->
<!-- ``` -->
*Math*. Mathematically, histogram can be produced for any (finite) sequence of
numbers: we divide the range into several bins, count how many of the points in
the sequence falls into each bin, and then draw a bar above that bin whose
height is equal (or proportional to) that count. The picture tells use about how
the sequence we started from is "distributed". The order of the points does not
matter - you would get exactly the same picture if you sorted the points first.
If the sequence of points you draw the histogram of comes from, say, normal
distribution, the histogram will resemble the shape of the pdf of a normal
distribution. I say resemble, because its shape is ultimately random. If the
number of points is small (like in the second part of this problem) the
histogram may look nothing like the normal pdf. However, when the number of
points gets larger and larger, the shape of the histogram gets closer and closer
to the underlying pdf (if it exists). I keep writing "shape" because the three
histograms above have very different scales on the $y$ axis. That is because we
used counts to set the vertical sizes of bins.
A more natural choice is to use the proportions, i.e. relative frequencies (i.e.
counts divided by the total number of points) for bar heights. More precisely, the bar height $h$ over the bin $[a,b]$ is chosen so that the area of the bar, i.e., $(b-a)\times h$ equals to the proportion of all points that fall inside $[a,b]$. This way, the total area under the histogram is always $1$. To draw such a **density histogram** in R we would
need to add the additional option `freq = FALSE` to `hist`:
```{r}
x = rnorm(100000, mean = 1, sd = 2)
hist(x,freq = FALSE)
```

Note how the $y$-axes label changed from "Frequency" to "Density". 
With such a normalization, the histogram of $x$ can be directly compared to the
probability density of a normal distribution. Here is a histogram of $100,000$
simulations from our normal distribution with its density function (pdf)
superimposed; I am leaving the code in case you are interested:

```{r echo=2:6}
set.seed(1098)
sims = rnorm(10000, mean = 1, sd = 2)
x = seq(-6,8,by=0.02)
y = dnorm(x, mean = 1, sd = 2)
hist(sims, freq=FALSE, main = "")
points(x,y, type="l", lwd=3, col="red" )
```

## Multivariate Distributions

```{exercise}
Let `x` contain $2,000$ draws from $N(0,1)$, `z` another $2,000$ draws from $N(0,1)$ and let `y=x^2+z`. 

1. Draw a scatterplot of `x` and `y` to visualize the joint distribution of `x` and `y`

2. Plot two histograms, one of `x` and one of `y`. Do they tell the whole story about the joint distribution of `x` and `y`?
   
3. Are `x` and `y` correlated? Do `x` and `y` in your plot "look independent"?  Use the permutation test to check of independence between `x` and `y`. 
```
**Solution:**

1.\  

```{r}
x = rnorm(2000)
z = rnorm(2000)
y = x^2+z
plot(x,y)
```
2. \ 

```{r echo=2:3}
par(mfrow=c(1,2))
hist(x)
hist(y)
```

No, the two histograms would not be enough to describe the joint distribution.
There are many ways in which two random variables $X$ and $Y$ can be jointly
distributed, but whose separate (marginal) distributions match the histograms above. To give a very simple example,
let $X$ and $Y$ be discrete random variables, each of which can only take values
$0$ or $1$. Consider the following two possible *joint* distribution tables for the random pair $(X,Y)$:


```{r echo=FALSE}
library(kableExtra)

jd1 = data.frame(c(0.25, 0.25),
                c(0.25, 0.25))
rownames(jd1) = c(" 0"," 1")
colnames(jd1) = c('0','1')

jd2 = data.frame(c(0.5, 0),
                c(0, 0.5))
rownames(jd2) = c(" 0"," 1")
colnames(jd2) = c('0','1')

t1 = kable(jd1) %>% 
   row_spec(0, align="center") %>% 
   column_spec(1, bold = T) %>% 
   kable_styling(full_width = F, "striped")

t2 = kable(jd2) %>% 
   row_spec(0, align="center") %>% 
   column_spec(1, bold = T) %>% 
   kable_styling(full_width = F, "striped") 

knitr::kables(list(t1,t2))

   
```

In both cases, the marginals are the same, i.e., both $X$ and $Y$ are equally likely to take the value $0$ or $1$, i.e., they both have the Bernoulli distribution with parameter $p=1/2$. That would correspond to the separate histograms to be the same. On the other hand, their joint distributions (aka dependence structures) are completely different. In the first (left) case, $X$ and $Y$ are independent, but in the second they are completely dependent.

3.\  They are probably not correlated since the sample correlation between `x` and `y` is close to $0$:
```{r}
(cor(x,y))
```
but they do not look independent.


To apply the permutation test, we first plot the scatterplot of `x` vs. `y` as above. Then, we replace `y` by a vector with the same components, but  randomly permute their positions, and then plot a scatterplot again. We repeat this three times:
```{r echo=2:12}
par(mfrow=c(2,2))
y_perm_1 = sample(y)
y_perm_2 = sample(y)
y_perm_3 = sample(y)
plot(x,y)
plot(x,y_perm_1)
plot(x,y_perm_2)
plot(x,y_perm_3)

```

The conclusion is clear, the first (upper-left) plot is very different than the other three. Therefore, `x` and `y` are probably not independent. 


**Comments.**

*Math.* The point of this problem is to review the notion of the **joint distribution** between two random variables. The most important point here is that there is more to the joint distribution of two random vectors, than just the two distributions taken separately. In a sense, the whole is (much) more than the sum of its parts. This is something that does not happen in the deterministic world. If you give me the $x$-coordinate of a point, and, separately, its $y$-coordinate, I will be able to pinpoint the exact location of that point. 

On the other hand, suppose that the $x$-coordinate of a point is unknown, so we treat it as a random variable, and suppose that this variable admits the standard normal distribution. Do the same for $y$. Even with this information, you cannot say anything about the position of the point $(x,y)$. It could be that the reason we are uncertain about $x$ and the reason we are uncertain about $y$ have nothing to do with each other; in that case we would be right to assume that $x$ and $y$ are independent. If, on the other hand, we got the values of  both $x$ and $y$ by measuring them using the same, inaccurate, tape measure, we cannot assume that the errors are independent. It is more likely that both $x$ and $y$ are too big, or both $x$ and $y$ are too small. 

 Mathematically, we say that random variables $X$ and $Y$ are independent if
$$ \PP[X \in [a,b]] \times \PP[ Y \in [c,d] ] = \PP[ X\in [a,b] \text{ and } Y\in [c,d]]\text{ for all } a,b,c,d.$$
While up to the point, this definition is not very eye-opening, or directly applicable in most cases. Intuitively, $X$ and $Y$ are independent if the distribution of $Y$ would not change if we received additional information about $X$. In our problem, random variables $X$ and $Y$ correspond to vectors `x` and `y`. Their scatterplot above clearly conveys the following message: when `x` is around $-2$, we expect `y` to be around `4`, while when `x` is around $0$, `y` would be expected to be around $0$, too. 

Sometimes, it is not so easy to decide whether two variables are independent by staring at a scatterplot. What would you say about the scatterplot below?
```{r echo=FALSE, fig.asp=1, out.width="50%"}
x1 = sin(runif(3000,0,2*pi)**2)
y1 = cos(runif(3000,0,2*pi)**2)
plot(y1~x1, xlab="x", ylab="y",asp=1, xlim=c(-1,1))
```
The **permutation test** is designed to help you decide when two (simulated) random variables are likely to be independent. The idea is simple. Suppose that `x` and `y` are simulations from two independent (not necessarily identical) distributions; say `x=runif(1000)` and `y=rnorm(1000)`. The vector `y_perm=sample(y)` is a randomly permuted version of `y` (see R section below) and it contains exactly the same information about the distribution of `y` as `y` itself does. Both `y` and `y_perm` will produce exactly the same histogram. Permuting `y`, however, "uncouples" it from `x`. If there was any dependence between the values of `x` and `y` before, there certainly isn't any now. In other the joint distribution of  `x` and `y_perm` has the same marginals as the joint distribution of `x` and `y`, but all the (possible) dependence has been removed. What remains is to compare the scatterplot between `x` and `y` and the scatterplot between `x` and `y_perm`. If they look about the same, we conclude that `x` and `y` are independent. Otherwise, there is some dependence between them. 

One question remains: why did we have to draw three scatterplots of permuted versions of `y`? That is because we have only finitely many data points, and it can happen, by pure chance, that the permutation we applied to `y` does not completely scramble its dependence on `x`. With a "sample" of three such plots, we 
get a better feeling for the inherent randomness in this permutation procedure, and it is much easier to tell whether "one of these things is not like the others". Btw, the random variables in the scatterplot above are, indeed, independent; here are the $4$ permutation-test plots to "prove" it:
```{r, fig.asp=1, echo=FALSE}
x1 = sin(runif(3000,0,2*pi)**2)
y1 = cos(runif(3000,0,2*pi)**2)
par(mfrow=c(2,2))
plot(y1~x1, xlab="x", ylab="y",asp=1, xlim=c(-1,1))
plot(sample(y1)~x1, xlab="x", ylab="y_perm_1",asp=1, xlim=c(-1,1))
plot(sample(y1)~x1, xlab="x", ylab="y_perm_2",asp=1, xlim=c(-1,1))
plot(sample(y1)~x1, xlab="x", ylab="y_perm_3",asp=1, xlim=c(-1,1))

```

Unlike univariate (one-variable) distributions which are visualized using histograms or similar plots, multivariate (several-variable) distributions are harder to depict. The most direct relative of the histogram is a **3d histogram**. Just like the $x$-axis is divided into bins in the univariate case, in the multivariate case we divide the $xy$-plane into regions (squares, e.g.) and count the number of points falling into each of these regions. After that a 3d bar (a skyscraper) is drawn above each square with the height of each skyscraper equal (or proportional)  to the number of points which fall into its base. Here is a 3d histogram of our original pair (`x`,`y`) from the problem. You should be able to rotate and zoom it right here in the notes, provided your browser has JavaScript enabled:
```{r, echo=FALSE}
htmltools::includeHTML("pics/test.html")
```

A visualization solution that requires less technology would start the same way, i.e., by dividing the $xy$ plane into regions, but instead of the third dimension, it would use different colors to represent the counts. Here is an example where the regions are hexagons, as opposed to squares; it just looks better, for some reason:

```{r echo=FALSE}
library(hexbin)
df=data.frame(x,y)
hexbinplot(y~x, data=df, colramp = heat.ob, asp=1)
```

Just to showcase the range of possibilities, here is another visualization technique which  which requires deeper statistical tools, namely the **density contour plot**:
```{r echo=FALSE, out.width="80%"}
library(ggplot2)
ggplot(data=df, aes(x=x, y=y) ) +
  geom_density_2d(color="black", bins=12)+
   theme_classic()+
   theme(aspect.ratio=0.8)
```

*R.* There is very little new R here. 
You should remember that if `x` and `y` are vectors of the same length,  `plot(x,y)` gives you a scatterplot of `x` and `y`. 

To compute the sample correlation between two vectors, use the `cor`. 

We used the command `sample(y)` to obtain a randomly permuted version of `y`. The simplicity of this is due to default parameters of the command `sample` which we already learned about. In particular, the default number of samples is exactly the size of the input vector `y` and, by default, sampling is performed *without replacement*. If you think about it for a second, you will realize that a sample of size $n$ from the vector of size $n$ *without* replacement is nothing by a random permutation of `y`. 

You are not required to do this in your submissions, but if you want to  display several plots side-by-side, use the command is `par(mfrow=c(m,n))` before the `plot` commands. It tells R to plot the next $mn$  plots in a $m\times n$ grid. 

```{exercise}
Random variables $X$ and $Y$ have the joint distribution given by the following table:
<table class="table table-striped" style="width: auto !important; margin-left: auto; margin-right: auto;">
 <thead>
  <tr>
   <th style="text-align:left;text-align: center;">   </th>
   <th style="text-align:right;text-align: center;"> 1 </th>
   <th style="text-align:right;text-align: center;"> 2 </th>
   <th style="text-align:right;text-align: center;"> 3 </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;font-weight: bold;"> 1 </td>
   <td style="text-align:right;"> 0.1 </td>
   <td style="text-align:right;"> 0.2 </td>
   <td style="text-align:right;"> 0.3 </td>
  </tr>
  <tr>
   <td style="text-align:left;font-weight: bold;"> 2 </td>
   <td style="text-align:right;"> 0.2 </td>
   <td style="text-align:right;"> 0.2 </td>
   <td style="text-align:right;"> 0.0 </td>
  </tr>
</tbody>
</table>
   
Simulate $10,000$ draws from the distribution of $(X,Y)$ and display a 
contingency table of your results.
```


## Additional Problems

```{exercise} 
<br>
   
1. Find the Weibull distribution in R's help system.<!-- ' -->
Simulate $n=10000$ draws from the Weibull distribution with shape parameter $2$ and scale parameter $3$. Draw a histogram of your simulations. 

2.  Suppose that the vector `x` contains $n=10000$ simulations from the standard
    normal $\mu=0, \sigma=1)$. Without simulating any new random numbers,
    transform it into the vector `y` such that `y` is a vector of $n=10000$
    simulations from the normal with $\mu=1$ and $\sigma=0.5$. Draw histograms
    of both `x` and `y` on the same plot. (*Note:* the extra parameter `add` is
    used to superimpose plots. You may want to use different colors, too. Use
    the e parameter `col` for that. )

3.  Starting with `x=seq(-3,3,by=0.1)`, define the appropriate vector `y` and
    use `x` and `y` to plot the graph of the cdf of the standard normal. The
    command you want to use is `plot` with the following extra arguments
   - `type="l"` (to get a smooth line instead of a bunch of points). 
   - `main="The CDF of the standard normal"` (to set the title), and
   -  another argument (which you must look up youself) that will set the $y$-axis label to $F(x)$. 
```

```{exercise}
<br>
   
1. Simulate $n=1000$ draws from the  distribution whose distribution table is given by
<table class="table" style="width: auto !important; margin-left: auto; margin-right: auto;">
 <thead>
  <tr>
   <th style="text-align:right;"> 2 </th>
   <th style="text-align:right;"> 4 </th>
   <th style="text-align:right;"> 8 </th>
   <th style="text-align:right;"> 16 </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:right;"> 0.2 </td>
   <td style="text-align:right;"> 0.3 </td>
   <td style="text-align:right;"> 0.1 </td>
   <td style="text-align:right;"> 0.4 </td>
  </tr>
</tbody>
</table>
and plot their histogram.

2.  You may have learned in probability how to compute the pdf $f_Y(y)$ of a
    transformation $Y=g(X)$ of a random variable with pdf $f_X(x)$. Suppose
    that you forgot how to do that, but have access to $10,000$ simulations from
    the distribution of $X$. How would you get an approximate idea about the
    shape of the function $f_Y$? <br>
   More concretely, take $X$ to be exponentially distributed with parameter $1$ and
   $g(x) = \sin(x)$ and produce a picture that approximates the pdf $f_Y$ of $Y$.
   (Note: even if you remember how to do this analytically, you will run into a
   difficulty. The function $\sin(x)$ is not one-to-one and the method usually
   taught in probability classes will not apply. If you learned how to do it in the
   many-to-one case of $g(x)= \sin(x)$, kudos to your instructor!)

3.  Let $X$ be a random variable with the Cauchy distribution, and $Y =
    \operatorname{arctan}(X)$. R allows you to simulate from the Cauchy
    distribution, even if you do not know what it is. How would you use that to
    make an educated guess as to what the distribution of $Y$ is? To make your
    life easier, consider $\tfrac{2}{\pi} Y$ first. 

```

```{exercise}
A basic method for obtaining simulations draws from distributions other than the
uniform is the **transformation method**. The idea is to start with (pseudo)
random numbers, i.e., draws from the uniform $U(0,1)$ distribution, and then
apply a function $g$ to each simulation. The difficulty is, of course, how to
choose the right function $g$.
<br>
Let $X$ be a random variable with a continuous and strictly increasing cdf $F$.
What is the distribution of $Y=F(X)$? What does that have to do with the transformation method?
```
 
```{exercise} 
(Extra credit) 

1. Let $f_1$ and $f_2$ be two pdfs. We take a constant $\alpha \in (0,1)$ and define the function $f$ by
$$ f(x) = \alpha f_1(x) + (1-\alpha) f_2(x).$$
The function $f$ is the pdf of a third distribution, which is called the
**mixture of $f_1$ and $f_2$ with weights $\alpha$ and $1-\alpha$**.
Assuming that you know how to simulate from the distributions with pdfs $f_1$
and $f_2$, how would you draw $10,000$ simulations from the mixture $f$? Show
your method on the example of a mixture of $N(0,1)$ and $N(4,1)$ with
$\alpha=2/3$. Plot the histogram of the obtained sample (play with the
parameter `breaks` until you get a nice picture.)
   
   (*Hint:* start with two vectors, the first containing $10,000$ simulations from
$f_1$ and the second from $f_2$. Then "toss" $10,000$ biased coins with
$\mathbb{P}[ H ] = \alpha$ ... )

2. The **double exponential** or
**Laplace** distribution is a continuous probability distribution whose pdf is
given by $$ \tfrac{1}{2} \exp(-|x|), x\in {\mathbb R}.$$ This distribution
is not built into R. How would you produce simulations from the double
exponential using R?
```

⬇︎ In case you were wondering, the text below belongs to footnotes from somewhere high above.⬇︎