\newcommand{\PP}{{\mathbb{P}}}
\newcommand{\RR}{{\mathbb{R}}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\N}{{\mathbb{N}}}
\newcommand{\sS}{{\mathcal{S}}}
\newcommand{\EE}{{\mathbb{E}}}
\newcommand{\ld}{\lambda}
\newcommand{\eand}{\text{ and }}

# (APPENDIX) Appendix {-}

# Probability Distributions {#dist}

Here are the basic facts about the probability distributions we will need in these lecture notes. For a much longer list of important distributions, check  this [wikipedia page](https://en.wikipedia.org/wiki/List_of_probability_distributions). 


## Discrete distributions:

Note: $(q=1-p)$
```{r echo=FALSE, warning = FALSE}
library(tidyverse)
library(kableExtra)
discrete = matrix(c(
      c("",
      "Parameters",
      "Notation",
      "Support",
      "pmf",
      "$\\EE[X]$",
      "$\\Var[X]$"
   ),
   c( "Bernoulli",
      "$p\\in (0,1)$", 
      "$B(p)$", 
      "$\\{0,1\\}$", 
      "$(q,p,0,0,\\dots)$", 
      "$p$", 
      "$pq$" 
      # "$ps+q$"
      ),
  c("Binomial",
      "$n\\in\\N , p\\in (0,1)$",
      "$b(n,p)$",
      "$\\{0,1,\\dots, n\\}$",
      "$\\binom{n}{k} p^k q^{n-k}$",
      "$np$",
      "$npq$"
      # "$(ps +q)^n$"
      ),
   c("Geometric",
      "$p\\in (0,1)$",
      "$g(p)$",
      "$\\{0,1,\\dots\\}$",
      "$p q^k$",
      "$q/p$",
      "$q/p^2$"
      # "$\\tfrac{p}{1-qs}$"
   ),
  c("Poisson",
      "$\\lambda\\in(0,\\infty)$",
      "$P(\\lambda)$",
      "$\\{0,1,\\dots\\}$",
      "$e^{-\\lambda} \\tfrac{\\lambda^k}{k!}$",
      "$\\lambda$",
      "$\\lambda$"
      # "$e^{\\lambda(s-1)}$"
   )),
  byrow = TRUE,
  ncol = 7
) 

a = as_tibble(discrete)

kable(a, col.names = NULL) %>% 
   column_spec(1, bold = TRUE) %>% 
   row_spec(1, italic = TRUE)

```

## Continuous distributions:
Note: the pdf is given by the formula in the table only on its support. It is equal to $0$ outside of it. 
```{r echo=FALSE, warning = FALSE}
library(tidyverse)
library(kableExtra)
continuous = matrix(c(
      c("",
      "Parameters",
      "Notation",
      "Support",
      "pdf",
      "$\\EE[X]$",
      "$\\Var[X]$"
   ),
   c( "Uniform",
      "$a\\lt b$",
      "$U(a,b)$", 
      "$(a,b)$", 
      "$\\frac{1}{b-a}$",
      "$\\frac{a+b}{2}$", 
      "$\\frac{(b-a)^2}{12}$" 
      ),
   c( "Normal",
      "$\\mu\\in{\\mathbb R},\\sigma \\gt 0$", 
      "$N(\\mu,\\sigma)$", 
      "${\\mathbb R}$", 
      "$\\frac{1}{\\sigma \\sqrt{2\\pi}} e^{-\\tfrac{(x-\\mu)^2}{2 \\sigma^2}}$", 
      "$\\mu$", 
      "$\\sigma^2$" 
      ),
   c( "Exponential",
      "$\\lambda\\gt 0$",
      "$\\operatorname{Exp}(\\lambda)$", 
      "$(0,\\infty)$", 
      "$\\lambda e^{-\\lambda x}$",
      "$\\tfrac{1}{\\lambda}$", 
      "$\\frac{1}{\\lambda^2}$"
      )
   ),
  byrow = TRUE,
  ncol = 7
)


a = as_tibble(continuous)

kable(a, col.names = NULL) %>% 
   column_spec(1, bold = TRUE) %>% 
   row_spec(1, italic = TRUE)

```